---
title: "Interlacer Example"
author: "Kyle Husmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Trying out {interlacer}

```{r}
library(tidyverse)
library(interlacer)

int_df <- read_interlaced_csv(
  "w1_mathproj_stu_svy_raw.csv",
  na = c("refused", "skipped")
)

na_levels(int_df$math3)

# See if the labelled missing are recognized in sum and friends (yes)

int_df %>%
  summarize(sum = sum(math3, na.rm = TRUE))

int_df %>%
  summarize(sum = mean(math3, na.rm = TRUE))

int_df %>%
  summarize(sum = sd(math3, na.rm = TRUE))

int_df %>%
  mutate(foo = math3 + math4)

int_df %>%
  rowwise() %>%
  mutate(
    mean_math = mean(c_across(starts_with("math")), na.rm=TRUE) 
  )

# Also recognized in filter

int_df %>%
  filter(is.na(math3))

# filter also uses labels by default

int_df %>%
  filter(math3 == na("skipped"))
```

Here's some things you can't do in labelled:

```{r}
# We can add missing values to types other than character and integer.
# interlacer works for ALL possible R types.

# For example, let's replace the dates for a given stu_id with a missing value:

(int_df_date <- int_df %>%
  mutate(
    svy_date = if_else(stu_id == 1347, na("technical error"), svy_date)
  ))

# Note that the missing reason channel AUTOMATICALLY adds "technical error"
# to its list of missing values:

na_levels(int_df_date$svy_date)

# Also notice that svy_date is still a date type

is.Date(int_df_date$svy_date)

# This means when we plot it, the axis will show up as a date. Notice the X axis!

int_df_date %>%
  ggplot(aes(svy_date, math1)) +
  geom_point()

# If we were using labelled, we'd have to turn everything into character vectors,
# and we'd lose all the date features!
```

Let's say now that we want to code our data to be compatible with SPSS.
We'll do two things:

1) Code gender to f = 1, m = 2, and
2) Code skipped = -99, refused = -98

```{r}
# Because values and missing reasons are separate channels, it's easy to do
# target them separately in tidy pipelines

(int_df_coded <- int_df %>%
  mutate(
    gender = map_value_channel(gender, \(v) recode(v, f = 1, m = 2)),
    across_na_channels(everything(), \(v) recode(v, skipped = -99, refused = -98)),
  )
)

# Notice that because the mutations occur *within* the value and missing reason
# channels, there's no need to set -99 and -98 as missing values after, as
# you would with labelled. They've just been transformed into numeric values.

int_df_coded %>%
  summarize(sum = sum(math3, na.rm = TRUE))

# We get the same result as before!

# Note that because this missing reason channel is now numeric, it doesn't
# have factor levels:
na_levels(int_df_coded$math3)
```

## Comparing to previous workflow with labelled

```{r}
library(labelled)

lab_df <- read_csv("w1_mathproj_stu_svy_raw.csv")
```

```{r, eval = FALSE}
# Can assign skipped as missing but this variable will remain character
# You receive an error when you try to convert to numeric

lab_df_coded <- lab_df %>%
  set_na_values(math3 = "skipped") %>%
  mutate(math3 = as.numeric(math3))
```

```{r}
# Instead you need to use numeric missing values (e.g., -99 = skipped)

lab_df_coded <- lab_df %>%
  mutate(
    # This is annoying -- you have to convert everything to character first, because
    # some math cols are dbl, and some are character
    across(everything(), as.character),
    # Also here when I recode, I have to use strings for the numeric values, because
    # I'm still working in characters
    across(everything(), \(v) recode(v, "skipped" = "-99", "refused" = "-98")),
    gender = recode(gender, f = "1", m = "2"),
    # Ok, now my coded columns are numeric
    across(c(starts_with("math"), "gender"), as.numeric),
    # But I have to convert my dates back to date types
    svy_date = as.Date(svy_date)
  ) %>%
  set_na_values( # Is there no way to do this with across()?
    math1 = c(-99, -98),
    math2 = c(-99, -98),
    math3 = c(-99, -98),
    math4 = c(-99, -98),
    gender = c(-99, -98),
  ) %>%
  set_value_labels( # Similarly, there's no across() compatibility here either?
    math1 = c(skipped = -99, refused = -98),
    math2 = c(skipped = -99, refused = -98),
    math3 = c(skipped = -99, refused = -98),
    math4 = c(skipped = -99, refused = -98),
    gender = c(f = 1, m = 2, skipped = -99, refused = -98)
  )

na_values(lab_df_coded$math3)

# But even then, the missing labelled values are not properly masked in computations:

lab_df_coded %>%
  summarize(sum = sum(math3, na.rm = TRUE))

lab_df_coded %>%
  summarize(sum = mean(math3, na.rm = TRUE))

lab_df_coded %>%
  summarize(sum = sd(math3, na.rm = TRUE))

lab_df_coded %>%
  mutate(foo = math3 + math4)

lab_df_coded %>%
  rowwise() %>%
  mutate(
    mean_math = mean(c_across(starts_with("math")), na.rm=TRUE) 
  )

# They are recognized in filter

lab_df_coded %>%
  filter(is.na(math3))

# But by default, filtering for specific missing values requires referencing
# their arbitrary
# codes

lab_df_coded %>%
  filter(math3 == -99)

# You can get the labels by converting to factor though!
lab_df_coded %>%
  filter(as_factor(math3) == "skipped")
```

Ok, let's write our coded data frame to a file, and load it with interlacer:

```{r}
write_csv(lab_df_coded, "w1_mathproj_stu_svy_coded.csv")

(int_df_coded <- read_interlaced_csv(
  "w1_mathproj_stu_svy_coded.csv",
  na = c(-99, -98)
))
```

Ok, this gives me a data frame with all my codes. But these are annoying to work with!
If I want to filter by gender, for example, I write:

```{r}
int_df_coded %>%
  filter(gender == 1)
```

Or by missing reason:

```{r}
int_df_coded %>%
  filter(gender == na(-98))
```

With interlacer, you can load a coded dataset with `cfactor` types instead which
allows you to work with labels by default, but then you have codes whenever you
need them:

```{r}
(int_df_coded_cfct <- read_interlaced_csv(
  "w1_mathproj_stu_svy_coded.csv",
  col_types = x_cols(
    gender = v_col_cfactor(codes = c(f = 1, m = 2))
  ),
  na = c(skipped = -99, refused = -98)
))
```

now we can filter with labels:

```{r}
int_df_coded_cfct %>%
  filter(gender == "f")

int_df_coded_cfct %>%
  filter(gender == na("refused"))
```

But we always can transform back into codes if we wish:

```{r}
# transform a single channel on a single column
int_df_coded_cfct |>
  mutate(
    gender = map_value_channel(gender, as.codes)
  )

# Or transform all of them
int_df_coded_cfct |>
  mutate(
    across_value_channels(where_value_channel(is.cfactor), as.codes),
    across_na_channels(where_na_channel(is.cfactor), as.codes),
  )
```

The goal with `cfactor` is to avoid magic numbers in our code by using labels
instead. This can help reduce errors and make our code much more readable.
